---
title: "Import"
author: "EB"
date: "April 14, 2019"
output: html_document
---

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)
```

## Reading different filetypes

This is a list of commands to make R read various types of datafiles

```{r}

library(readr)

# Reading CSV file directly
mydata1 <- read.csv("c:/data/mydatafile.csv",header = TRUE)
# Open window to search for *.csv
mydata2 <- read.csv(file.choose(), header = TRUE)
#Read Semi-colon Delimited files
mydata3 <- read.csv2("file2.csv", header = TRUE)
#Read files with any delim
mydata4 <- read.delim("file.txt", delim = "|")
#Read fixed width file
mydata5 <- read.fwf("file.xxx", col_positions = c(1, 3, 5))
#Tab delim files
mydata6 <- read.table("file.xxx")
#Read Excel files
mydata7 <- readxl("file.xlsx")
```

## Exploring data as it is

A few most basic functions that I always forget. Allow you to look into your data and understand its structure

```{r}
# Provides the structure of the dataset
str(mydata)
# Provides basic descriptive statistics and frequencies
summary(mydata)
# Lists variables in the dataset
names(mydata)
# First 6 rows of dataset
head(mydata)
# First 10 rows of dataset. If (n = -10) then all rows but last 10
head(mydata, n=10)
# Last 6 rows, same trick with (n = X) applies
tail(mydata) 
# First 10 rows of data
mydata[1:10, ]
# First 10 rows of data of the first 3 variables - like 2D matrix
mydata[1:10,1:3]
# View raw data - seems to require dplyr library
View(mydata)
# Sort dataframe by column Y in descending order
mydata[order(mydata$Y), decreasing = TRUE]
# Number of rows in the dataset
nrow(mydata)
# Number of columns in the dataset
ncol(mydata)

```


## DPLYR

Dplyr is a versatile package that allows you to do analysis quickly. It is useful for creating tables of summary statistics across specific groups of data. 


```{r}

library(dplyr)

# Goto template for using dplyr
mydata %>%   # Original dataframe to work with. Funny %>% passes object on the left as argument of a function on the right
  filter(X > 30) %>%  # Applying filter 
  group_by(Y, Z) %>%  # Grouping variables
  summarise(
    a = mean(col.a),       # calculate mean of column col.a in my.df
    b = sd(col.b),         # calculate sd of column col.b in my.df
    c = max(col.c))        # calculate max on column col.c in my.df, ...

```


