---
title: "Wrangling"
author: "EB"
date: "April 14, 2019"
output: html_document
---

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)
```

## Tidy Data

There are three rules to ensure a dataset is tidy:
1. Each variable must have its own column.
2. Each observation must have its own row.
3. Each value must have its own cell.


In addition to my goto dplyr template (see Data_Import), in the tidyverse most of the functions are defined as action verbs that you can apply to data.

- `filter()` - remove rows according to a criteria
- `select()` - select columns by name
- `mutate()` - calculate new column variables by manipulating data
- `arrange()` - sort data by columns

Tibbles are like frame class, but have three advantages:
1. Subsetting - `[` always returns a new tibble, `[[` and `$` always return a vector.
2. No partial matching - You must use full column names when subsetting
3. Display - When you print a tibble, R provides a concise view of the data that fits on one screen


```{r}
library(tidyverse)

mytibble <- as_tibble(mydata) # Convert data frame to tibble
mytibble <- tbl_df(mydata) # Seemingly alternative way to make a tibble
gather(mytibble, "colname1", "colname2", 2:4) # Gather columns into rows (e.g., for datasets where columns are variables). Opposite - spread()
separate(mytibble, date, c("y", "m", "d")) # Separate one column into several - useful with dates. Opposite - unite()

distinct(mytibble) # Remove duplicate rows
slice(mytibble, 10:15) # Select rows by position.
select(mytibble, colname1, colname2, colname3) #Select columns by name or helper function - contains(), ends_with(), matches(regex), start_with(), etc.

drop_na(mytibble) # Drop rows containing NA's
fill(mytibble, .direction = c("down", "up")) # Fill in NA’s with most recent values.
replace_na(mytibble, list(x = 5)) # Replace NA’s by column

```

### Summarize data

Basically, `summarise` boils down data to a single row of value. `summarise` uses summary functions, such as:
  first - First value of a vector;
  last - Last value of a vector;
  nth - Nth value of a vector;
  n - Number of values in a vector;
  n_distinct - Number of distinct values in a vector;
  IQR - Interquartille range of a vector;
  min/max - Minimum/maximum value in a vector;
  mean/median/var/sd - Mean, median, variance, standard deviation of a vector.


```{r}
summarise(mytibble, avg = mean(colname1))
```

### New variables

`mutate` Applies functions and/or adds more new columns. `mutate` uses functions that convert one vector of variables into another, such as:
  lead/lag - Copy with values shifted/lagged by 1;
  row_number - Ranks. Ties got to first value;
  between - Checks if values are between x and y;
  cummean/cumsum/cummax/cummin - Cumulative mean/sum/max/min.


```{r}
mutate(mytibble, new_col = X + Y)
```
